Flooding of the Okavango Delta influences Connectivity for Dispersing African Wild Dogs

### 2022-09-13 Chapter_8/04_Manuscript/Manuscript.tex
Major update. Adjusted generation of source areas.

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
Climate change is expected to considerably impact vital rates of endangered
species. In particular dispersal, the phase during which species leave their
natal area and attempt to find a new territory to settle, may

We utilize seasonal changes in the flooding regime across the Okavango delta as
a large scale natural experiment to investigate how environmental changes impact
species' ability to disperse, and utlimately connectivity between remaining
subpopulations. Specifically, we apply individual-based dispersal simualtions to
simulate dispersal trajectories under two extreme scenarios, assuming minimal
and maximum flood extents.

Our results show that during years of minimum flood extent, the Okavango delta
releases vast areas that dispersers use as movement corridors, whereas during
years of extreme flood the delta's floodwaters pose a considerable dispersal
barrier that forces dispersing individuals closer towards human settlements.
Besides a better understanding of the conservation needs for the African wild
dog, our study also provides evidence that incorporating seasonality in studies
of connectivity is imperative to more accurately predict the dispersal ability
of endangered species.

@@

\section{Introduction}
\subsection{Climate Change}
\subsection{Vital Rates}
\subsection{Dispersal}
\subsection{Connectivity}
Dispersal is an important, if not the important, driver of landscape
connectivity and therefore of major interest to conservation authorities.

\subsection{Okavango Delta and African Wild Dogs}
The Okavango delta in Southern Africa poses a unique opportunity to study the
impacts of environmetal change on species dispersal ability and connectivity in
large scale natural experiment setup.

\section{Methods}
Programming language, packages, etc.

\subsection{Study Area}
The study area for this analysis was focused on the Okavango delta and its
surroundings in Southern Africa, comprising parts of Angola, Namibia, Botswana,
Zimbabwe, and Zambia \Cref{StudyArea}. To accomodate for the long distance
dispersal events commonly observed in the African wild dog, we considered a
large rectangular extent stretching from ... to ...., totalling to an area of xx
km. The dominant geographical feature in this region is the Okavango delta, the
world's largest inland delta and main driver of seasonal environmental change in
the region. While local rainfalls marginally impact water levels across the
delta, the main flood-pulsing rythm is driven by the water-influx from the
Okavango river. The Okavango carries rain waters collected the Angolian
highlands into the delta's distributaries. Although precipitation in Angola
peaks around ..., the water arrives in the delta with a delay of ... It only
slowly descends through the delta and reaches the distal ends (at the faults)
around ... At minimum extent, the flood covers an area of around xx km, whereas
during maximum flood it covers ... Vegetation in the study area can largely be
categorized into dense mopane forest, mixed acacia woodland, and plain
grassland. Human influence in the study area is generally low and mainly
concentrated to small villages at the western periphery of the delta, as well as
to the city of Maun at the southern tip of the delta. Large portions of land are
dedicated national parks, game reserves or forest reserves and the area is part
of the world's largest transboundary conservation initiative, the
Kavango-Zambezi Transfrontier Conservation Area.

@@

\subsection{Connectivity}
Based on simulated dispersal trajectories we quantified connectivity across the
OD for both environmental scenaiors in three complementary way. First, we
generated heatmaps showing how often different areas in the landscape were
visited by simulated dispersers. Such heatmaps serve to detect dispersal
hotspots, yet are less suitable to detect pinchpoints and bottlenecks. Hence, we
also computed spatially explicit betweenness scores which are useful to
highlight movement corridors (Bastille). For this, we overlayed the study area
with a regular grid with 2.5 km x 2.5 km grid cells and determined how often
simulated individuals transitioned from one grid-cell to another. Note that we
didn't consider revisits to avoid pronouncing regions in which individuals were
encroached and moved in circles.

To gain insights into landscape connectivity from the simulated dispersal
events, we prepared three complementary connectivity maps (see again
\citep{Hofmann.2022}); a heatmap, indicating the absolute frequency at which
each pixel in the study area was traversed by a virtual disperser, a betweenness
map, highlighting pixels of particular importance to facilitate movement from
one area to another, and a map of interpatch connectivity, depicting the
presence and intensity of functional links between source areas. To compute the
heatmap, we counted how many times each pixel in the study area was traversed by
a virtual disperser. To calculate betweenness, we overlayed the study area with
a regular grid and used the simulated trajectories to determine movements from
one grid-cell to another. Specifically, we counted how often transitions from
one grid-cell to any other occured. The centerpoint of each grid-cell then
served as network node, and the number of transitions from one cell to another
as weights between the respective nodes. Based on the so generated network, we
then computed betweenness scores for all network nodes (i.e. for all grid-cells
across the study area). Lastly, to compute interpatch connectivity, we
determined the frequency at which virtual dispersers originating from one source
area reached any of the other source areas. We also calculated the average
dispersal duration needed until individuals reached the respective area.

\section{Results}

@@

\section{Discussion}

>>>>

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
Climate change is expected to profoundly impact vital rates of wild living
animal populations. While the impact of climate change on the demographics of
local subpopulations has been studied repeatedly, little is known about the
consequences of environmental change on dispersal and connectivity between such
subpopulations.

Here, we capitalize on natural, flood-pulse driven environmental change across
the Okavando Deltas in northern Botswana to investigate the impact of changing
environmental conditions on the dispersal ability of the endangered African wild
dog. For this, we utilize a previously parametrized dispersal model and apply it
to simulate dispersal under two extreme environmental scenarios; one assuming
minimal flood extent, one assuming maximum flood extent.

In genral, our results suggest only a marginal impact of the flood extent on
dispersal prospects. Even during maximum flood, dispersal between neighboring
patches remained high. Between a few selected patches, however, dispersal was
substantially restricted during times of high flood. That is the Okavango delta
posed a substantial dispersal barrier that reduced connectivity between patches
during maximum flood, but revealed vital dispersal corridors that facilitated
dispersal into neighboring areas during minimum flood. Besides a better
understanding of the conservation needs for the African wild dog, our study also
presents a first step towards incorporating environmental change due to
seasonality or climate change. It also emphasizes the usefulness of
individual-based dispersal simulations as a conservation tool to study impacts
of environmental change in a spatially realistic way.

@@

\section{Introduction}
\subsection{Climate Change}

Despite the importance of climate change in determining species viability,
predicting the way and extent to which it will affect species remains
challenging. One exciting avenue to gain insights into the implications of
environmental change is to utilize seasonal variation in the environment.

Cite IPCC, Planetary boundaries, Tucker (step length). Phenological shifts. Cite
Robynne, Arpat, Megan, Dominik, Rabaiotti etc.

\subsection{Vital Rates}
Several studies investigating the impacts of climate change on life-history
traits have revealed that population persistence may critically depends on
climate conditions \citep{Paniw.2019}. While many studies examine climate
impacts on survival and reproductive rates, the influence of global change or
climatic extreme events on dispersal prospects has received little attention.

Ecosystems that undergoe seasonal environmental variation, in particular arid
regions that experience extreme climatic conditions, are among the most
vulnerable to climate change.

\subsection{Dispersal}
\subsection{Connectivity}
Dispersal is an important, if not the important, driver of landscape
connectivity and therefore of major interest to conservation authorities.

\subsection{Okavango Delta and African Wild Dogs}
The Okavango delta in Southern Africa poses a unique opportunity to study the
impacts of environmetal change on species dispersal ability and connectivity in
large scale natural experiment setup. According to projections by the IPCC,
Botswana and its surroundings are among the most vulnerable to climate change.

\section{Methods}
We conducted all analyses using the programming language R.

\subsection{Study Area}
The study area for this analysis was focused on the Okavango delta and its
surroundings in Southern Africa, comprising parts of Angola, Namibia, Botswana,
Zimbabwe, and Zambia (\Cref{StudyArea}). To accommodate for the long distance
dispersal events commonly observed in African wild dogs (e.g.
\citep{DaviesMostert.2012, Masenga.2016, Cozzi.2020}), we considered a large
rectangular extent stretching from 20\degree 30' E to 26\degree E. totaling to
an area of 300'000 km\textsuperscript{2}. The dominant geographical feature in
this region is the Okavango delta (OD), the world's largest inland delta and the
main driver of seasonal environmental change in the region. The flood-pulsing
rhythm of the OD is mainly dictated by precipitation in the Angolian highlands,
where rainfalls are channeled into the OD through the Okavango River. Although
precipitation in Angola peaks in ..., water only slowly descends through the
Okavango river and its tributaries, reaching the distal ends (i.e. the faults)
of the delta in July or August. At minimum extent, the flood covers an area of
3'600 km\textsuperscript{2}, whereas during maximum flood it covers more than
9'000 km\textsuperscript{2}. Vegetation in the study area is dominated by dense
mopane forest, mixed acacia woodland, and plain grassland. Human influence in
the study area is low and mainly concentrated around small villages at the
western periphery of the delta and to the city of Maun at the southern tip of
the delta. Large portions of land are dedicated national parks, game reserves or
forest reserves. The study area is also part of the world's largest
transboundary conservation initiative, the Kavango-Zambezi Transfrontier
Conservation Area.

@@

\subsection{Connectivity}
Based on simulated dispersal trajectories in the two scenarios we quantified
connectivity in three complementary ways. First, we generated heatmaps depicting
the frequency at which different areas in the landscape were visited by
simulated dispersers. Such heatmaps serve to detect dispersal hotspots but are
less suitable to detect pinchpoints and bottlenecks. Hence, we also computed
spatially explicit betweenness scores which are useful to highlight such
pinchpoints (Bastille). To compute betweenness, we overlayed the study area with
a regular grid with 2.5 km x 2.5 km grid cells and determined how often
simulated individuals transitioned from one grid-cell to another. In cases where
the same individual repeatedly realized the same cell-transition, we only
counted a single transition to avoid emphasizing regions regions in which
individuals moved in circles. Finally, we calculated the number of successful
dispersal events between the different source areas. We coin this type of
connectivity ``inter-patch connectivity''. Dispersal between two areas was said
to be successful whenever a trajectory leaving one area intersected with the
target area. To gauge the dispersal duration needed to move between patches (be
consistent with ``patches'', ``source-areas'' etc.), we also recorded the
minimum number of steps that individuals had to move to move between areas.

\section{Results}
Heatmaps produced from simulated dispersal trajectories reveal that the OD acts
as a major dispersal barrier during periods of high flood, but reveals viable
dispersal corridors during periods of low flood. In particular, our analysis
reveals that the area north-west of Maun is frequently visited by dispersers at
in the minimum-flood scenario, yet that an extended flood, coupled with high
human influence around the city of Maun, enforces simulated individuals to
either detour around Maun, or to slip through the narrow window between village
and flood. Similarly, the betweenness map indicates the presence of a dispersal
corridor extending from the western section of the delta leading across the
delta's center into its western part, yet only during low-flood. During maximum
flood, the same corridor is pushed south and enforces dispersers to move closer
to the city of Maun. In general, connectivity during maxium flood is lower
compared to the minimum flood scenario (number links vs. number liks). Most
notably, source area 5 is almost inaccessible during maximum flood.

 Most notably, the
area north-west of Maun acts as dispersal habitat during phases of low flood,
yet is almost inaccessible during phases of maxium flood. This is reflected in
the heatmap, where connectivity across the western part of the delta is low
during maximum flood, but also in the betweenness map, where an important
movement corridors through the delta is only visible during minimum flood.
Although connectivity across large parts of the landscape remains unchanged
across environmental scenarios, one critical movement corridor shifts in
response to an expanding flood. Most notably, a corridor that links the western
and eastern part of the delta during minimum flood gets redirected further
south, thus leading more closely to the city of Maun. Unsurprisingly, the
differences are more pronounced for areas directly affected by the flood, yet it
is worth noting that even some areas not directly affected by the flood.

@@

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_Betweenness.png}
  \caption{}
  \label{Betweenness}
  \end{center}
\end{figure}

\input{IPC_Frequency.tex}
\input{IPC_Duration.tex}

\section{Discussion}
According to our simulations, the propensity to move between the eastern and
western part of the delta is much lower during maximum extent. This is mainly
due to the flood-waters and the city of Maun acting as dispersal barriers.
During maximum extent, the floodwaters of the delta close a gap between the
delta and Maun that otherwise would serve as dispersal corridor. Anecdotal
evidence supports this hypothesis, for the only dispersing individuals recorded
to move from the eastern to the western part of the delta moved at times of low
flood. In line with this, it appears that a large flood extent pushes dispersing
individuals to move closer to human inhabited areas such as the village of Maun.

Predicting how climate change will impact the dispersal ability of AWDs is
challenging for multiple reasons. First of all, predicting the flooding patterns
of the OD under climate change is merely impossible due to the complex feedbacks
between surface-temperature, soil conditions, precipitation patterns and the
associated changes in vegetation. Second, the delta is not only prone to changes
in environmental conditions, but also to changes in anthropogenic use of the
inflowing water. Finally, it is unclear how AWDs, in fact, how any
species, will cope with environmental change due to global warming. Even though
some studies predict that AWD populations are likely to decline under increasing
temperatures, these studies fail to account for the behavioral plasticity of
their focal species. AWDs respiratory system, for instance, has evolved as a
perfect adaptation to high temperatures and AWDs may, in fact, profit from a
comparative advantage (cite an economist) over their competitors and prey under
rising temperatures. Although the theory of comparative advantages is a
fundamental concept in economics, it has yet to find its way into ecological
studies.

\cite{MurrayHudson.2006} predicted that increased temperatures, additional human
abstractions, and reduced river flows might lead to a ``Delta dying'' and that
the impact of climate will be much more pronounced than the impact of
anthropogenic water use.

Although local rainfalls in Botswana are expected to increase in terms of
intensity, simulations show that the length of the rainy season will decline,
more than offsetting the incline in precipitation \cite{Akinyemi.2019}.

According to our simulations, dispersers are able to cover larger distances
during periods of low flood. This finding is little surprising, considering that
inundated areas act as dispersal barriers and force dispersers to detour and
circumvent water-covered areas. However, it still leads to an interesting
hypothesis. Previous studies have shown that the euclidean dispersal distance of
female coalitions is larger than that of male coalitions. This has led to ...
However, demographic analyses have also revealed that female offspring tend to
emigrate from their pack at younger ages and earlier in the year, when
floodwaters are still at a relatively low level (Behr). It is thus conceivable
that the sex-differences in reported dispersal distances is mainly a consequence
of environmental conditions during dispersal, rather than owed to physiological
differences between sexes.

While our analysis marks an important step into incorporating environmental
change into studies of connectivity, there are several critical additions that
should be considered in the future. We studied dispersal and connectivity under
two different environmental scenarios, yet our movement model assumed that
dispersers had identical habitat and movement preferences in both scenarios. In
reality, however, it can be expected that movement and habitat kernels of
dispersers differ depending on the season considered (examples).

An additional complication arises when species movement is not solely driven by
environmental conditions, but also affected by itra- and inter-specific factors.
For instance, ... has shown that dispersers...
Rendering such conditions alone is challenging, yet rendering the conditions
under changing environmental conditions is merely impossible.

To address such differences, researchers could model
habitat and movement preferences using season-dependent models, or,
alternatively, by combining hidden markov models with step-selection functions.
(cite papers that fieberg sent)

Only recently, it has been discovered that AWDs communicate using shared marking
sites. The role of such marking sites for dispersing coalitions remains to be
investigated, yet it is is likely that, akin to resident packs, use SMS as
navigation waypoint and demarcation lines. Chemical analyses suggest that the
compounds used for communication are highly volatile and may not persist in
extreme climate conditions. In result, dispersers may lose their ability to
effectively navigate across the landscape and locate potential mates with whom
to settle. This would reduce pack-formation prospects and undermine...

Validating predictions from individual-based dispersal models is challenging and
requires additional dispersal data, which is inherently scarce anyways. Sciticen
science may help to fill this gap by augmenting observed GPS data with
occasional sightings of uncollared dispersing coalitions. This is especially
ciritical for species that disperse across borders and beyond confined study
areas. The African carnivore wildbook offers ...

The OD is an important driver of species distribution and it has been found that
an expanding flood limits available habitat, thus leading to more inter-specific
competition, particularly between AWDs and lions.

\subsection{Conclusion}
Our dispersal simulations across two extreme climatic scenarios reveal striking
differences in dispersal prospects and landscape connectivity for dispersing
AWDs. This implies that (1) climatic variation, be it due to seasonality or
climate change, must be included in anlayses dealing with dispersal and (2) that
projected climate change is likely to have profound impacts on landscape
connectivity.
